---
title: "CS5812 Predictive Data Analysis"
author: "Mohammad Ahmad Author (STUDENT_ID)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Libraries

```{r}
rm(list = ls())
gc()

library(ggplot2)
library(rpart)
```

## 2. Load dataset

```{r}
df <- read.csv("final_london_modelling.csv", stringsAsFactors = FALSE)

dim(df)
names(df)
head(df)
```

## 3. Initial checks

```{r}
str(df)
summary(df)
```

## 4. Remove duplicate transaction IDs if present

```{r}
if ("transaction_id" %in% names(df)) {
  before_rows <- nrow(df)
  df <- df[!duplicated(df$transaction_id), ]
  after_rows <- nrow(df)
  
  cat("Rows before duplicate removal:", before_rows, "\n")
  cat("Rows after duplicate removal :", after_rows, "\n")
  cat("Rows removed                 :", before_rows - after_rows, "\n")
} else {
  cat("transaction_id column not found. Duplicate transaction check skipped.\n")
}
```

## 5. Drop unused columns

```{r}
drop_cols <- c(
  "X",
  "X.1",
  "manhattan_station_km",
  "dist_to_tube_km",
  "n_exact_keys",
  "locality",
  "postcode",
  "transaction_id",
  "street_split_clean"
)

df <- df[, !(names(df) %in% drop_cols)]

dim(df)
names(df)
```

## 6. Create target variable

```{r}
df$price <- as.numeric(df$price)

df <- df[!is.na(df$price) & df$price > 0, ]

df$log_price <- log(df$price)

summary(df$price)
summary(df$log_price)
```

## 7. Date and time variables

```{r}
if ("date_of_transfer" %in% names(df)) {
  df$date_of_transfer <- as.Date(df$date_of_transfer, format = "%d-%m-%Y")
  
  df$sale_year <- as.numeric(format(df$date_of_transfer, "%Y"))
  df$sale_month <- as.numeric(format(df$date_of_transfer, "%m"))
}

cat("Sale year range :", min(df$sale_year, na.rm = TRUE), "to", max(df$sale_year, na.rm = TRUE), "\n")
cat("Sale month range:", min(df$sale_month, na.rm = TRUE), "to", max(df$sale_month, na.rm = TRUE), "\n")

if (max(df$sale_year, na.rm = TRUE) < 1900) {
  stop("Date parsing error: sale_year is not a real year. Check date_of_transfer format.")
}

summary(df$sale_year)
summary(df$sale_month)
```

## 8. Convert categorical variables

```{r}
factor_cols <- c(
  "built_form",
  "construction_age_band",
  "floor_level",
  "old_new",
  "duration",
  "district",
  "epc_source",
  "epc_timing_flag",
  "property_type",
  "station_zone"
)

for (col in factor_cols) {
  if (col %in% names(df)) {
    df[[col]] <- as.factor(df[[col]])
  }
}

str(df)
```

## 9. Select modelling variables

```{r}
model_cols <- c(
  "price",
  "log_price",
  "current_energy_efficiency",
  "potential_energy_efficiency",
  "built_form",
  "total_floor_area",
  "number_habitable_rooms",
  "construction_age_band",
  "floor_level",
  "heating_cost_current",
  "heating_cost_potential",
  "hot_water_cost_current",
  "hot_water_cost_potential",
  "lighting_cost_current",
  "old_new",
  "duration",
  "district",
  "epc_source",
  "epc_timing_flag",
  "lat",
  "long",
  "straight_line_station_km",
  "station_zone",
  "property_type",
  "sale_year",
  "sale_month"
)

model_cols <- model_cols[model_cols %in% names(df)]

df_model <- df[, model_cols]

dim(df_model)
names(df_model)
```

## 10. Missing-value summary

```{r}
missing_summary <- data.frame(
  column = names(df_model),
  missing_count = colSums(is.na(df_model)),
  missing_percent = round(colSums(is.na(df_model)) / nrow(df_model) * 100, 2)
)

missing_summary <- missing_summary[order(-missing_summary$missing_count), ]

missing_summary
```

## 11. Remove remaining missing rows

```{r}
before_rows <- nrow(df_model)

df_model <- na.omit(df_model)

after_rows <- nrow(df_model)

cat("Rows before removing missing values:", before_rows, "\n")
cat("Rows after removing missing values :", after_rows, "\n")
cat("Rows removed                       :", before_rows - after_rows, "\n")
```

## 12. Output folders and plot theme

```{r}
dir.create("outputs", showWarnings = FALSE)
dir.create("eda_plots", showWarnings = FALSE)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
)

set.seed(42)
sample_size <- min(10000, nrow(df_model))
plot_sample <- df_model[sample(1:nrow(df_model), sample_size), ]
```

## 13. Exploratory Data Analysis

### 13.1 Distribution of log price

```{r}
p1 <- ggplot(df_model, aes(x = log_price)) +
  geom_histogram(bins = 50, fill = "#4C78A8", color = "white") +
  labs(
    title = "Distribution of log price",
    x = "Log price",
    y = "Count"
  )

p1
ggsave("eda_plots/r_eda_log_price_distribution.png", p1, width = 8, height = 5, dpi = 300)
```

### 13.2 Distribution of raw price

```{r}
p2 <- ggplot(df_model, aes(x = price)) +
  geom_histogram(bins = 50, fill = "#F58518", color = "white") +
  labs(
    title = "Distribution of price",
    x = "Price",
    y = "Count"
  )

p2
ggsave("eda_plots/r_eda_price_distribution.png", p2, width = 8, height = 5, dpi = 300)
```

### 13.3 Property type and log price

```{r}
p3 <- ggplot(df_model, aes(x = property_type, y = log_price, fill = property_type)) +
  geom_boxplot(show.legend = FALSE) +
  labs(
    title = "Log price by property type",
    x = "Property type",
    y = "Log price"
  )

p3
ggsave("eda_plots/r_eda_property_type_boxplot.png", p3, width = 8, height = 5, dpi = 300)
```

### 13.4 District and log price

```{r}
district_counts <- sort(table(df_model$district), decreasing = TRUE)
top_districts <- names(district_counts)[1:15]

df_top_districts <- df_model[df_model$district %in% top_districts, ]

p4 <- ggplot(df_top_districts, aes(x = district, y = log_price, fill = district)) +
  geom_boxplot(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Log price by top districts",
    x = "District",
    y = "Log price"
  )

p4
ggsave("eda_plots/r_eda_district_boxplot.png", p4, width = 8, height = 6, dpi = 300)
```

### 13.5 Floor area and log price

```{r}
p5 <- ggplot(plot_sample, aes(x = total_floor_area, y = log_price)) +
  geom_point(alpha = 0.25, color = "#2E86AB") +
  geom_smooth(method = "lm", se = FALSE, color = "#D62728") +
  labs(
    title = "Floor area and log price",
    x = "Total floor area",
    y = "Log price"
  )

p5
ggsave("eda_plots/r_eda_floor_area_log_price.png", p5, width = 8, height = 5, dpi = 300)
```

### 13.6 Habitable rooms and log price

```{r}
p6 <- ggplot(plot_sample, aes(x = number_habitable_rooms, y = log_price)) +
  geom_point(alpha = 0.25, color = "#1F9D8A") +
  geom_smooth(method = "lm", se = FALSE, color = "#D62728") +
  labs(
    title = "Rooms and log price",
    x = "Number of habitable rooms",
    y = "Log price"
  )

p6
ggsave("eda_plots/r_eda_rooms_log_price.png", p6, width = 8, height = 5, dpi = 300)
```

### 13.7 Straight-line station distance and log price

```{r}
p7 <- ggplot(plot_sample, aes(x = straight_line_station_km, y = log_price)) +
  geom_point(alpha = 0.25, color = "#7A5195") +
  geom_smooth(method = "lm", se = FALSE, color = "#D62728") +
  labs(
    title = "Station distance and log price",
    x = "Straight-line distance to station (km)",
    y = "Log price"
  )

p7
ggsave("eda_plots/r_eda_station_distance_log_price.png", p7, width = 8, height = 5, dpi = 300)
```

### 13.8 Energy efficiency and log price

```{r}
p8 <- ggplot(plot_sample, aes(x = current_energy_efficiency, y = log_price)) +
  geom_point(alpha = 0.25, color = "#E45756") +
  geom_smooth(method = "lm", se = FALSE, color = "#2E2E2E") +
  labs(
    title = "Energy efficiency and log price",
    x = "Current energy efficiency",
    y = "Log price"
  )

p8
ggsave("eda_plots/r_eda_energy_efficiency_log_price.png", p8, width = 8, height = 5, dpi = 300)
```

## 14. Correlation analysis

```{r}
numeric_cols <- c(
  "price",
  "log_price",
  "current_energy_efficiency",
  "potential_energy_efficiency",
  "total_floor_area",
  "number_habitable_rooms",
  "heating_cost_current",
  "heating_cost_potential",
  "hot_water_cost_current",
  "hot_water_cost_potential",
  "lighting_cost_current",
  "lat",
  "long",
  "straight_line_station_km"
  "sale_year",
  "sale_month"
)

numeric_cols <- numeric_cols[numeric_cols %in% names(df_model)]

numeric_vars <- df_model[, numeric_cols]

cor_matrix <- cor(numeric_vars, use = "complete.obs")

round(cor_matrix, 2)
```

## 15. Correlation heatmap

```{r}
cor_df <- as.data.frame(as.table(cor_matrix))
names(cor_df) <- c("variable_1", "variable_2", "correlation")

p9 <- ggplot(cor_df, aes(x = variable_1, y = variable_2, fill = correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#3B4CC0",
    mid = "white",
    high = "#B40426",
    midpoint = 0
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Correlation heatmap",
    x = "",
    y = ""
  )

p9
ggsave("eda_plots/r_correlation_heatmap.png", p9, width = 9, height = 7, dpi = 300)
```

## 16. K-means clustering for exploratory analysis

```{r}
set.seed(42)

cluster_data <- df_model[, c("lat", "long", "straight_line_station_km", "log_price")]

cluster_sample <- cluster_data[
  sample(1:nrow(cluster_data), min(5000, nrow(cluster_data))),
]

cluster_scaled <- scale(cluster_sample[, c("lat", "long", "straight_line_station_km")])
```

### 16.1 Elbow method

```{r}
wcss <- numeric(10)

for (k in 1:10) {
  set.seed(42)
  wcss[k] <- kmeans(cluster_scaled, centers = k, nstart = 20)$tot.withinss
}

elbow_df <- data.frame(
  clusters = 1:10,
  wcss = wcss
)

p10 <- ggplot(elbow_df, aes(x = clusters, y = wcss)) +
  geom_line(color = "#4C78A8") +
  geom_point(color = "#F58518", size = 2.5) +
  labs(
    title = "Elbow method for k-means",
    x = "Number of clusters",
    y = "Within-cluster sum of squares"
  )

p10
ggsave("eda_plots/r_kmeans_elbow.png", p10, width = 8, height = 5, dpi = 300)
```

### 16.2 K-means with 3 clusters

```{r}
set.seed(42)

kmeans_result <- kmeans(cluster_scaled, centers = 3, nstart = 20)

table(kmeans_result$cluster)

cluster_sample$cluster <- as.factor(kmeans_result$cluster)

p11 <- ggplot(cluster_sample, aes(x = long, y = lat, color = cluster)) +
  geom_point(alpha = 0.6, size = 1.4) +
  labs(
    title = "K-means clusters",
    x = "Longitude",
    y = "Latitude",
    color = "Cluster"
  )

p11
ggsave("eda_plots/r_kmeans_clusters.png", p11, width = 8, height = 5, dpi = 300)
```

### 16.3 Cluster price summary

```{r}
cluster_summary <- aggregate(
  log_price ~ cluster,
  data = cluster_sample,
  FUN = median
)

names(cluster_summary) <- c("cluster", "median_log_price")

cluster_summary

write.csv(cluster_summary, "outputs/r_kmeans_cluster_summary.csv", row.names = FALSE)
```

## 17. Train-test split

```{r}
set.seed(42)

train_index <- sample(1:nrow(df_model), 0.8 * nrow(df_model))

train_data <- df_model[train_index, ]
test_data <- df_model[-train_index, ]

cat("Training rows:", nrow(train_data), "\n")
cat("Testing rows :", nrow(test_data), "\n")
```

## 18. Decision Tree Regression

```{r}
predictor_names <- setdiff(names(train_data), c("price", "log_price"))

formula_dt <- as.formula(
  paste("log_price ~", paste(predictor_names, collapse = " + "))
)

formula_dt
```

```{r}
dt_model <- rpart(
  formula_dt,
  data = train_data,
  method = "anova",
  control = rpart.control(
    cp = 0.001,
    maxdepth = 8,
    minsplit = 50
  )
)

printcp(dt_model)
```

## 19. Decision Tree plot

```{r}
png("eda_plots/r_decision_tree_plot.png", width = 1200, height = 800)
plot(dt_model, uniform = TRUE, margin = 0.1)
text(dt_model, use.n = TRUE, cex = 0.6)
dev.off()

plot(dt_model, uniform = TRUE, margin = 0.1)
text(dt_model, use.n = TRUE, cex = 0.6)
```

## 20. Predictions

```{r}
predicted_log <- predict(dt_model, newdata = test_data)
actual_log <- test_data$log_price
```

## 21. Evaluation functions

```{r}
rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}

mae <- function(actual, predicted) {
  mean(abs(actual - predicted))
}

r2_score <- function(actual, predicted) {
  1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2)
}
```

## 22. Decision Tree evaluation on log scale

```{r}
dt_rmse_log <- rmse(actual_log, predicted_log)
dt_mae_log <- mae(actual_log, predicted_log)
dt_r2_log <- r2_score(actual_log, predicted_log)

cat("Decision Tree - Log Price Results\n")
cat("RMSE:", round(dt_rmse_log, 4), "\n")
cat("MAE :", round(dt_mae_log, 4), "\n")
cat("R2  :", round(dt_r2_log, 4), "\n")
```

## 23. Decision Tree evaluation on original price scale

```{r}
actual_price <- exp(actual_log)
predicted_price <- exp(predicted_log)

dt_rmse_price <- rmse(actual_price, predicted_price)
dt_mae_price <- mae(actual_price, predicted_price)
dt_r2_price <- r2_score(actual_price, predicted_price)

cat("Decision Tree - Price Results\n")
cat("RMSE:", round(dt_rmse_price, 2), "\n")
cat("MAE :", round(dt_mae_price, 2), "\n")
cat("R2  :", round(dt_r2_price, 4), "\n")
```

## 24. Median baseline model

```{r}
baseline_log_prediction <- median(train_data$log_price)

baseline_predicted_log <- rep(baseline_log_prediction, length(actual_log))

baseline_rmse_log <- rmse(actual_log, baseline_predicted_log)
baseline_mae_log <- mae(actual_log, baseline_predicted_log)
baseline_r2_log <- r2_score(actual_log, baseline_predicted_log)

baseline_actual_price <- exp(actual_log)
baseline_predicted_price <- exp(baseline_predicted_log)

baseline_rmse_price <- rmse(baseline_actual_price, baseline_predicted_price)
baseline_mae_price <- mae(baseline_actual_price, baseline_predicted_price)
baseline_r2_price <- r2_score(baseline_actual_price, baseline_predicted_price)

cat("Median Baseline - Price Results\n")
cat("RMSE:", round(baseline_rmse_price, 2), "\n")
cat("MAE :", round(baseline_mae_price, 2), "\n")
cat("R2  :", round(baseline_r2_price, 4), "\n")
```

## 25. Actual vs predicted plot

```{r}
results_df <- data.frame(
  actual_log = actual_log,
  predicted_log = predicted_log,
  actual_price = actual_price,
  predicted_price = predicted_price,
  residual = actual_log - predicted_log
)

set.seed(42)
plot_results <- results_df[
  sample(1:nrow(results_df), min(10000, nrow(results_df))),
]

p12 <- ggplot(plot_results, aes(x = actual_log, y = predicted_log)) +
  geom_point(alpha = 0.25, color = "#4C78A8") +
  geom_abline(slope = 1, intercept = 0, color = "#D62728") +
  labs(
    title = "Decision Tree: actual vs predicted log price",
    x = "Actual log price",
    y = "Predicted log price"
  )

p12
ggsave("eda_plots/r_decision_tree_actual_vs_predicted.png", p12, width = 8, height = 5, dpi = 300)
```

## 26. Residual plot

```{r}
p13 <- ggplot(plot_results, aes(x = predicted_log, y = residual)) +
  geom_point(alpha = 0.2, color = "#7A5195") +
  geom_hline(yintercept = 0, color = "#D62728") +
  labs(
    title = "Decision Tree residual plot",
    x = "Predicted log price",
    y = "Residual"
  )

p13
ggsave("eda_plots/r_decision_tree_residuals.png", p13, width = 8, height = 5, dpi = 300)
```

## 27. Variable importance

```{r}
importance_df <- data.frame(
  feature = names(dt_model$variable.importance),
  importance = as.numeric(dt_model$variable.importance)
)

importance_df <- importance_df[order(-importance_df$importance), ]

importance_df
```

```{r}
top_importance <- head(importance_df, 15)

p14 <- ggplot(top_importance, aes(x = reorder(feature, importance), y = importance)) +
  geom_col(fill = "#54A24B") +
  coord_flip() +
  labs(
    title = "Top Decision Tree variables",
    x = "Feature",
    y = "Importance"
  )

p14
ggsave("eda_plots/r_decision_tree_importance.png", p14, width = 8, height = 5, dpi = 300)
```

## 28. Save output files

```{r}
test_predictions <- test_data
test_predictions$actual_log <- actual_log
test_predictions$predicted_log <- predicted_log
test_predictions$actual_price <- actual_price
test_predictions$predicted_price <- predicted_price
test_predictions$residual <- actual_log - predicted_log

metrics_df <- data.frame(
  model = "Decision Tree",
  rmse_log = dt_rmse_log,
  mae_log = dt_mae_log,
  r2_log = dt_r2_log,
  rmse_price = dt_rmse_price,
  mae_price = dt_mae_price,
  r2_price = dt_r2_price
)

baseline_metrics_df <- data.frame(
  model = "Median Baseline",
  rmse_log = baseline_rmse_log,
  mae_log = baseline_mae_log,
  r2_log = baseline_r2_log,
  rmse_price = baseline_rmse_price,
  mae_price = baseline_mae_price,
  r2_price = baseline_r2_price
)

comparison_metrics_df <- rbind(baseline_metrics_df, metrics_df)

write.csv(missing_summary, "outputs/r_missing_summary.csv", row.names = FALSE)
write.csv(results_df, "outputs/r_decision_tree_predictions.csv", row.names = FALSE)
write.csv(test_predictions, "outputs/r_decision_tree_test_predictions.csv", row.names = FALSE)
write.csv(importance_df, "outputs/r_decision_tree_importance.csv", row.names = FALSE)
write.csv(metrics_df, "outputs/r_decision_tree_metrics.csv", row.names = FALSE)
write.csv(baseline_metrics_df, "outputs/r_baseline_metrics.csv", row.names = FALSE)
write.csv(comparison_metrics_df, "outputs/r_model_baseline_comparison.csv", row.names = FALSE)

write.csv(train_data, "outputs/train_data_for_python.csv", row.names = FALSE)
write.csv(test_data, "outputs/test_data_for_python.csv", row.names = FALSE)

cat("R workflow completed successfully.\n")
cat("Train/test files exported for Python.\n")
```
